---
title: 'How much taxonomic expertise is necessary for prioritising areas of lowland rainforest sites as habitats for lichens in northeast Borneo?'
author: 'Rob Mrowicki'
date: '`r format(Sys.time(), "%d %B %Y")`'
output:
  pdf_document:
    fig_caption: yes
    fig_crop: yes
---

```{r init, include = FALSE}
source('./scripts/00_init.R')  # run initialisation script
```

```{r source_scripts, include = FALSE}
# run data scripts:
source('./scripts/01_import_tidy_and_manipulate_data.R')
source('./scripts/02_summarise_data.R')
source('./scripts/03_analyse_data.R')
source('./scripts/04_plotting.R')
```

## Field sites

Sampling was conducted at three sites within different areas of lowland forest in northeast Borneo, including two old growth reserve sites, (1) Danum Valley Conservation Area ('D') and (2) Maliau Conservation Area ('M'), and one selectively logged site, (3) the Stability of Altered Forest Ecosystems (SAFE) Project area ('S').

At each site, ... (D and M = 96; S = 72).

The following functional traits were recorded for each tree: (1) girth, measured as ...; (2) bark type (S = smooth, C = scaly, R = ridged/rough, DR = dry ancient); and (3) the presence of buttresses or flutes.

(Check actual numbers of replicates per site.)

```{r nobs_tables, eval = FALSE, echo = FALSE}
# insert tables for numbers of observations vs. site:
# ~ lichen taxonomic groups:
kable(nobs_site_li_taxa, caption = 'Lichen taxonomic group data')
# ~ lichen functional groups:
kable(nobs_site_li_func, caption = 'Lichen functional group data')
# ~ tree functional traits:
kable(nobs_site_tr_func, caption = 'Tree functional trait data')
```

Are there missing observations? (See _Queries_ below.)

## Data analysis

Taxonomic/functional group richness (_S_) and Shannon-Wiener diversity (_H\'_) were calculated for each tree.

Tree bark type was coded numerically in relation to increasing roughness, i.e. smooth = 0, scaly = 1, ridged/rough = 2, ancient dry = 3. (See _Queries_ below.)

Permutational analysis of variance (PERANOVA) was used to test for differences in lichen taxonomic and functional group richness and diversity among the three sites. Differences in multivariate lichen community structure among sites were tested using permutational multivariate analysis of variance (PERMANOVA), separately for taxonomic and functional groups. Analyses involved 9,999 permutations of residuals under a reduced model, and tests were based on Type II Sums of Squares owing to unequal numbers of observations among groups. (Normally, non-metric multidimensional scaling (MDS) plots would be used to visualise overall differences in community structure... however, MDS analyses failed to converge on a result.)

The biota--environment (BIOENV; Clarke & Ainsworth, 1993, _Mar. Ecol. Prog. Ser._ 92:205-219) routine was used to identify the optimal subset of tree functional traits (= 'environmental' variables) accounting for variability in lichen community structure, via maximisation of the rank correlation between environmental and biological distance matrices. The variables identified by the BIOENV analysis were then used in a canonical analysis of principal coordinates (CAP; Anderson & Willis, 2003, _Ecology_ 84:511--525) to enable dissimilarities in lichen communities to be visualised in relation to tree functional traits.

For all multivariate tests, community dissimilarities were calculated using the zero-adjusted Bray-Curtis coefficient (Clarke _et al._, 2006, _J. Exp. Mar. Biol. Ecol._ 330:55--80), based on $log10(x+1)$-transformed abundance data to reduce the influence of numerically dominant taxonomic/functional groups.

## Preliminary results

Mean lichen diversity (_H\'_ and _S_) per site:

```{r mean_div_tables, echo = FALSE}
# insert summary tables for mean lichen diversity vs. site:
# ~ taxonomic groups:
kable(mean_div_li_taxa, caption = 'Taxonomic group data')
# ~ functional groups:
kable(mean_div_li_func, caption = 'Functional group data')
```

![Boxplot of lichen taxonomic group richness (_S_) and diversity (_H\'_) versus site (D = Danum Valley Conservation Area, M = Maliau Conservation Area, S = Stability of Altered Forest Ecosystems (SAFE) Project area).](./figs/boxplots_taxa.pdf)

![Boxplot of lichen functional group richness and diversity versus site.](./figs/boxplots_func.pdf)

As expected, lichen richness and diversity (in terms of both taxonomic and functional groups) appeared to be lower at the logged site compared to the other two sites. While ANOVAs indicated significant differences in all cases, these results must be interpreted with caution owing to the poor fit of data to models -- specifically, assumptions of normality and/or homoscedasticity were contravened. (Perhaps because the data are zero-inflated, i.e. contain an excessively large number of empty observations? In this case, permutational ANOVA (PERANOVA) may be more suitable for univariate analyses.)

ANOVA results:

```{r anova_tables, echo = FALSE}
# insert ANOVA tables
# ~ taxonomic groups:
# ~~ richness:
kable(anova_li_taxa_s, caption = 'Lichen taxonomic group _S_ vs. site')
# ~~ diversity:
kable(anova_li_taxa_h, caption = 'Lichen taxonomic group _H\'_ vs. site')
# ~ functional groups:
# ~~ richness:
kable(anova_li_func_s, caption = 'Lichen functional group _S_ vs. site')
# ~~ diversity:
kable(anova_li_func_h, caption = 'Lichen functional group _H\'_ vs. site')
```

There were significant differences in overall lichen community structure (in terms of both taxonomic and functional groups) not only between old growth and logged sites, but also between Danum and Maliau Conservation Areas. (The particular taxonomic/functional groups making the greatest contributions to these differences may be examined in more detail via SIMPER analysis.)

PERMANOVA tables and post-hoc pairwise test results:

```{r permanova_tables, echo = FALSE}
# insert PERMANOVA tables and post-hoc pairwise test results:
# ~ taxonomic groups:
kable(
  perm_li_taxa,
  caption = 'Lichen community structure (taxonomic groups) vs. site'
)
kable(perm_li_taxa_ph, caption = 'Post-hoc pairwise tests')
# ~ functional groups:
kable(
  perm_li_func,
  caption = 'Lichen community structure (functional groups) vs. site'
)
kable(perm_li_func_ph, caption = 'Post-hoc pairwise tests')
```

Tree girth and presence of buttresses/flutes were identified by BIOENV as key tree traits influencing lichen community structure, in terms of both taxonomic and functional groups. CAP axis 1 correlated positively with tree girth, while axis 2 correlated negatively with buttress presence; however, only tree girth appeared to have a statistically significant effect on lichen communities.

(Relationships between lichen community structure and tree functional traits are to be visualised with a CAP ordination plot. Subsequent analysis will allow more detailed examination of which lichen taxonomic/functional groups correlate with particular tree traits.)

## Queries

1. In both lichen abundance datasets, there are two _D810_ columns -- is this a mistake (should these be combined into a single column in each case)? OR should one of these be labelled _D710_, as this tree is not present in either dataset?

1. I'm assuming that 'tree no.' is coded as follows: site-plot-tree (i.e. _D111_ = tree 11 in plot 1 in site Danum). Is this correct?

1. Can 'bark type' be coded numerically (i.e. an ordinal categorical variable), e.g. smooth = 1, scaly = 2, ridged/rough = 3, dry ancient = 4? In this way, it can be used as an environmental variable in the CAP analysis.
