---
title: 'How much taxonomic expertise is necessary for prioritising areas of lowland rainforest sites as habitats for lichens in northeast Borneo?'
author: 'Rob Mrowicki'
date: '`r format(Sys.time(), "%d %B %Y")`'
output:
  pdf_document:
    fig_caption: yes
    fig_crop: yes
  word_document:
    fig_caption: yes
---

```{r init, include = FALSE}
source('./scripts/00_init.R')  # run initialisation script
supp <- FALSE  # should 'supplementary' (i.e. 'tree-level') analyses be run?
```

```{r source_scripts, include = FALSE}
# run data scripts:
source('./scripts/01_import_tidy_and_manipulate_data.R')
source('./scripts/02_summarise_data.R')
source('./scripts/03_analyse_data.R')
source('./scripts/04_plotting.R')
```


## Field sites

Sampling was conducted at three sites within different areas of lowland forest in northeast Borneo, including two old growth reserve sites, (1) Danum Valley Conservation Area ('D') and (2) Maliau Conservation Area ('M'), and one selectively logged site, (3) the Stability of Altered Forest Ecosystems (SAFE) Project area ('S').

At each site, $\dots$ (number of trees: D and M = 96, S = 72; number of plots: D and M = 8, S = 6).

Lichens were categorised according to both (1) taxonomic and (2) functional group. Further, for each functional group, lichen functional traits were classified as follows: $\dots$ (Table _x_).

The following functional traits were recorded for each tree: (1) girth, measured by $\dots$; (2) bark type (DR = dry ancient, S = smooth, C = scaly, R = ridged/rough); (3) the presence of buttresses or flutes; and (4) dipterocarp vs. non-dipterocarp species. Additionally, for the majority of trees, (5) bark pH was determined from the mean of up to three readings, measured by $\dots$.

Tree girth and mean pH were converted into categorical variables (girth: s < 100 cm; m = 100--199 cm; l $\geq$ 200 cm; pH: vl < 2; l = 2 $\leq x <$ 4; m = 4 $\leq x <$ 6; h $\geq$ 6).^[These cut-off values for pH categories are quite arbitrary, and may need adjusting. Perhaps they could be based on the overall distribution of mean pH values, e.g. quartiles.]

For trees missing pH data, mean pH was estimated as the overall mean pH for the corresponding bark type category, justified by the expected relationship between bark type and pH,^[There doesn't appear to be a strong relationship between bark type and pH in this case. Also, modelling one tree trait based on another tree trait seems a bit circular, and may introduce unwanted covariance among 'environmental' variables in constrained multivariate analyses.], and then categorised accordingly.

```{r nobs_tables, eval = FALSE, echo = FALSE}
# check actual numbers of replicates per site
# combined table of numbers of replicates vs. site for each dataset:
list(
  rename(nobs_site_li_taxa, 'tax_grps' = 'n'),
  rename(nobs_site_li_func, 'func_grps' = 'n'),
  rename(nobs_site_tr_func, 'tree_traits' = 'n')
) %>%
  reduce(left_join, by = "site") %>%
  kable(caption = 'Number of replicates per site')
```


## Data analyses

Data analyses were conducted at 'plot-level', whereby plots, rather than individual trees, were treated as replicates.^[Increasingly, I felt that 'tree-level' analyses should be excluded altogether, with analyses focussing solely on 'plot-level' calculations, given that, for tree-level data, (1) some distance-based FD indices cannot be calculated and (2) MDS fails to converge (see below); also, (3) there was some concern over the non-independence of data when considering individual trees within the same plot as replicates, and (4) in practical terms, environmental assessments would be conducted at the scale of plots rather than individual trees. But perhaps tree-level results could still be included as Supplementary Material.] Thus, lichen taxonomic and functional group abundance data were summed for each plot;^[It might be worth checking whether exclusion of rare taxonomic/functional groups (e.g. overall mean abundance $\leq$ 0.1, or presence in <15% of plots) alters results substantially.] additionally, to enable plot-level characterisation of tree traits as continuous variables, we calculated the Shannon-Wiener diversity of (1) bark type and (2) pH categories,^[This approach seems a bit convoluted for pH, which is already a continuous variable. Alternatively, we could simply calculate the mean pH per plot, or perhaps the proportion of high/low pH trees (e.g. >6.5 or <4.5).] and the proportion of (3) large trees (i.e. girth $\geq$ 200 cm), (4) trees with buttresses or flutes and (5) dipterocarp trees within each plot.

Lichen taxonomic group^[It makes sense to calculate the three univariate diversity measures (richness, evenness and diversity) for taxonomic groups only, and not functional groups, with 'functional diversity' represented by the four distance-based FD indices.] richness (_S_), Simpson's evenness ($1-\lambda$) and Shannon-Wiener diversity (_H\'_) were calculated for each plot. Additionally, the following distance-based functional diversity indices were calculated from lichen functional group abundance data in conjunction with the functional traits matrix (Table _x_), using the function `dbFD` in the R package `FD`: functional richness (FRic), functional evenness (FEve), functional divergence (FDiv) and functional dispersion (FDis).^[Although all four distance-based FD indices were calculated successfully for each plot based on summed abundances, calculation of FRic and FDiv fails when using tree-level data, apparently owing to instances of <3 singular functional groups per tree. Further, it was not possible to calculate FEve and FDis for all individual trees, for similar reasons relating to data insufficiency.]

The community-level weighted means (CWMs) of lichen functional traits were also calculated for each plot, based on the summed abundances of trait classes (i.e. the product of the functional group abundance and functional trait classification matrices; see e.g. Garnier _et al._, 2004, _Ecology_ 85, 2630--2637; Lavorel _et al._, 2008, _Funct. Ecol._ 22:134--147; Koch _et al._, 2013, _Ecol. Indic._ 34:22--30), via the `dbFD::functcomp` function.

Prior to multivariate analyses (PERMANOVA, SIMPER, BIOENV and dbRDA -- see below),^[For some reason, $\log_{10}(x+1)$-transforming lichen abundance data prior to CAP site classification analyses tends to produce warnings resulting from negative Eigenvalues, and reduces overall classification success. Therefore these analyses are currently based on untransformed data.] lichen taxonomic and functional group abundance data were $\log_{10}(x+1)$-transformed to reduce the influence of numerically dominant taxonomic or functional groups.


### Differences in lichen diversity, community structure and traits among sites

Permutational analysis of variance (PerANOVA)^[Probability distributions are not really suitable for GLMMs, hence the use of PerANOVA. Another alternative is to use zero-inflated models, but  this introduces additional complications.] was used to test for differences in univariate lichen diversity measures (i.e. taxonomic richness, evenness and diversity, and distance-based functional diversity indices) among the three sites.

Differences in multivariate lichen community structure and trait composition among sites were tested using permutational multivariate analysis of variance (PERMANOVA), separately for taxonomic and functional groups, as well as for trait CWMs. Models included the factor 'site' (fixed, 3 levels: D, M and S), with a planned comparison (D, M) vs. S (i.e. old growth sites vs. logged site).^[For previous tree-level analyses, 'plot' (nested within 'site') was also included as a random factor to account for potential non-independence of trees located within the same plot.] Where a significant effect of 'site' was identified, pairwise permutational post-hoc tests were used to reveal differences between individual factor levels.^[It is probably better to use either the _a priori_ planned comparisons or post-hoc pairwise tests, and not both. The former is more appropriate if the aim of this study is to compare old growth vs. logged sites explicitly. But, more generally, adhering to 'old growth vs. logged' comparisons throughout would result in highly unbalanced analyses (i.e. 16 vs. 6 plots, or 192 vs. 72 trees).] Analyses involved 9,999 permutations of residuals under a reduced model, and tests were based on Type II Sums of Squares to account for unequal numbers of observations among groups.

To complement these unconstrained multivariate analyses, overall patterns in lichen community structure were visualised using non-metric multidimensional scaling (MDS) ordination plots, again based on both taxonomic and functional group abundances in addition to trait CWMs.^[The previous tree-level MDS analyses (performed using the `metaMDS` function in the R package `vegan`) failed to 'converge' on a result and produced relatively high 2D stress values, they may still serve as a useful representation of overall patterns -- to be interpreted with some caution. On the other hand, MDS converges successfully for summed plot-level abundance data.]

Relative contributions of individual lichen taxonomic and functional groups and trait classes to differences in community structure between groups of sites identified by PERMANOVA were subsequently determined via similarity of percentages (SIMPER; Clarke, 1993, _Aus. J. Ecol._ 18:117--143).

Univariate PerANOVAs were based on Euclidean distance matrices calculated from untransformed data. For all multivariate tests (both unconstrained and constrained, including dbRDA and CAP -- see below) community dissimilarities were calculated using the zero-adjusted Bray-Curtis coefficient (Clarke _et al._, 2006, _J. Exp. Mar. Biol. Ecol._ 330:55--80).^[The use of a 'zero-adjusted' dissimilarity coefficient is necessary for tree-level abundance data, which include a number of 'blank' samples; but we could probably just use standard Bray-Curtis for plot-level data.]


### Associations of lichen taxa, functional groups and trait classes with sites

The extent to which lichen taxonomic and functional groups, as well as functional traits, were associated with different combinations of sites was examined using extended Indicator Value analysis (De Cáceres _et al._, 2010, _Oikos_ 119:1674--1684), via the `multipatt` function in the R package `indicspecies` (v1.7.9; De Cáceres & Legendre, 2009). Analyses were based on the 'IndVal' index (Dufrêne & Legendre, 1997, _Ecol. Monogr._ 67:345--366), with the significance of associations tested using 9,999 permutations.


### Relationships of lichen communities and traits with tree functional traits

To enable dissimilarities in lichen communities and traits to be visualised in relation to tree functional traits, the five tree traits (= 'environmental' variables) were used in a constrained distance-based redundancy analysis (dbRDA), a variant of canonical analysis of principal coordinates (CAP; Anderson & Willis, 2003, _Ecology_ 84:511--525), using the `capscale` function in the R package `vegan` (version 2.5-6; Oksanen et al., 2019).^[While traditional CAP may be performed as a discriminant analysis (as opposed to a canonical correlation analysis) involving a single grouping variable, dbRDA allows for the inclusion of multiple categorical constraints.]

This analysis was performed separately for lichen taxonomic and functional groups and trait classes, using summed lichen group abundances, trait CWMs and continuous tree trait variables calculated for each plot as described above.^[If conducting these analyses at 'tree-level', to account for variation among plots in the tree-level analysis, 'plot' may be included as a conditioning variable in the dbRDA model.]

Prior to analyses, pairwise Spearman rank correlations were used to assess the extent of collinearity among tree traits. There were no strong pairwise correlations between bark type and pH diversity and proportions of large trees, trees with buttresses/flutes and dipterocarp trees ($|\rho| \leq\ 0.70$), therefore all five variables were included in subsequent analyses. The biota--environment (BIOENV; Clarke & Ainsworth, 1993, _Mar. Ecol. Prog. Ser._ 92:205--219) routine was used to identify the optimal subset of environmental variables accounting for variability in lichen community structure, via maximisation of the rank correlation between environmental and biological distance matrices. The variables identified by the BIOENV analysis were then used in the dbRDA analysis.^[I'm wondering whether or not to remove this part of the procedure, and simply include all 'environmental' variables in the constrained analysis.] Euclidean distances were calculated from environmental data, which were _z_-standardised to account for scaling differences between variables. The resulting ordination plots incorporated vectors representing Spearman rank correlations between continuous environmental variables and the first two CAP axes.

The significance of overall models (based on the sum of all eigenvalues) and of constraining variables (marginal terms) were assessed via ANOVA-like permutation tests involving `r format(n_perm, big.mark = ",")` permutations. Spearman rank correlation coefficients between individual taxonomic or functional group abundances and CAP axes were calculated to identify the most important groups and trait classes (i.e. $|\rho| \geq$ `r cor_spp_x`) contributing to variability in lichen communities.


### Classification of sites based on lichen communities and functional traits

Further CAP analyses were used to estimate the accuracy with which plots may be assigned to sites based on lichen community structure and trait composition. This cross-validation method involves the 'leave-one-out' approach of Lachenbruch & Mickey (1968; _Technometrics_ 10:1--11) to determine the proportion of individual observations (in this case, plots) that are successfully allocated to the correct group (Anderson & Willis, 2003). Using the `CAPdiscrim` function in the R package `BiodiversityR`,^[Previously, these analyses were conducted using the CAP FORTRAN programme (Anderson, 2002, _Ecological Archives_ E084-011-S1).] this analysis was performed separately for lichen taxonomic and functional groups and trait CWMs, versus a single factor, 'site' (i.e. a discriminant analysis, rather than a canonical correlation analysis). The appropriate numbers of principal coordinate axes (_m_) involved in the discriminant analyses were selected automatically based on the minimum misclassification error.


## Main results

Mean 'plot-level' lichen diversity (taxonomic group _S_, $1-\lambda$ and _H\'_; functional group FRic, FEve, FDiv and FDis) per site:

```{r mean_div_tables_plot, echo = FALSE}
# insert summary tables for mean lichen diversity vs. site:

# ~ taxonomic groups:
mean_div_li_taxa_plot %>%
  mutate_at(-1, ~ formatC(., digits = 2, format = "f")) %>%
  kable(caption = 'Taxonomic group data')

# ~ functional groups:
mean_div_li_func_plot %>%
  mutate_at(-1, ~ formatC(., digits = 2, format = "f")) %>%
  kable(caption = 'Functional group data')
```

![Boxplot of lichen taxonomic group richness (_S_), evenness ($1-\lambda$) and diversity (_H\'_) versus site (D = Danum Valley Conservation Area, M = Maliau Conservation Area {old growth sites}, S = Stability of Altered Forest Ecosystems (SAFE) Project area {logged site}).](./figs/boxplots_taxa_plot.pdf)

![Boxplot of distance-based Functional Diversity indices (FRic, FEve, FDiv and FDis) calculated for lichen functional groups versus site.](./figs/boxplots_func_plot.pdf)

Lichen taxonomic diversity and evenness were significantly lower at the logged site (S) compared to the two old growth sites (D and M), which in turn did not differ from each other (Figure _x_). For richness, although there was only a marginally significant effect of site (P `r rep_p(as.numeric(li_taxa_plot_s[li_taxa_plot_s$Source=="site", "P(MC)"]))`) and post-hoc pairwise comparisons revealed no significant difference between M and S, the planned comparison between logged and old growth sites suggested that richness was greater at the latter (`r rep_p(as.numeric(li_taxa_plot_s[li_taxa_plot_s$Source=="(D,M) vs. S", "P(MC)"]))`).

PerANOVA tables and post-hoc pairwise test results (taxonomic groups):

```{r peranova_tables_taxa_plot_s, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks
# insert PerANOVA tables:

# ~ taxonomic groups:
# ~~ richness:
li_taxa_plot_s %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen taxonomic group _S_ vs. site')

li_taxa_plot_s_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

```{r `peranova_tables_taxa_plot_1-l`, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# ~~ diversity:
`li_taxa_plot_1-l` %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen taxonomic group $1-\\lambda$ vs. site')

`li_taxa_plot_1-l_ph` %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

```{r peranova_tables_taxa_plot_h, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# ~~ diversity:
li_taxa_plot_h %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen taxonomic group _H\'_ vs. site')

li_taxa_plot_h_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

None of the four distance-based Functional Diversity indices showed clear differences between logged and old growth sites. Only for FDis was there a significant effect of site, whereby planned comparisons indicated a lower FDis at logged vs. old growth sites (P `r rep_p(as.numeric(li_func_plot_fdis[li_func_plot_fdis$Source=="(D,M) vs. S", "P(MC)"]))`); however, post-hoc pairwise comparisons were inconclusive. Although planned comparisons suggested that FRic was lower in the logged site compared to the old growth sites (P `r rep_p(as.numeric(li_func_plot_fric[li_func_plot_fric$Source=="(D,M) vs. S", "P(MC)"]))`), post-hoc comparisons were again inconclusive.^[One disadvantage of using 'plot-level' data for analyses is a loss of statistical power compared to when individual trees are treated as replicates.]

PerANOVA tables and post-hoc pairwise test results (functional groups):

```{r peranova_tables_func_plot_fric, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# ~ functional groups:
# ~~ FRic:
li_func_plot_fric %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen functional group FRic vs. site')

li_func_plot_fric_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

```{r peranova_tables_func_plot_feve, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# ~~ FEve:
li_func_plot_feve %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen functional group FEve vs. site')

li_func_plot_feve_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

```{r peranova_tables_func_plot_fdiv, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# ~~ FDiv:
li_func_plot_fdiv %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen functional group FDiv vs. site')

li_func_plot_fdiv_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

```{r peranova_tables_func_plot_fdis, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# ~~ FDis:
li_func_plot_fdis %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen functional group FDis vs. site')

li_func_plot_fdis_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

There were highly significant differences in overall lichen community structure (in terms of both taxonomic and functional groups) and functional trait composition (as trait CWMs) between old growth and logged sites, but not between Danum and Maliau Conservation Areas (Tables _x_). This pattern was evident from MDS plots showing strong separation between the SAFE Project area and the other two sites (Figure _x_).

PERMANOVA tables and post-hoc pairwise test results (taxonomic groups):

```{r permanova_tables_taxa_plot, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# insert PERMANOVA tables and post-hoc pairwise test results:
# ~ taxonomic groups:
perm_li_taxa_plot %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen community structure (taxonomic groups) vs. site')

perm_li_taxa_plot_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

PERMANOVA tables and post-hoc pairwise test results (functional groups):

```{r permanova_tables_func_plot, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# ~ functional groups:
perm_li_func_plot %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen community structure (functional groups) vs. site')

perm_li_func_plot_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

PERMANOVA tables and post-hoc pairwise test results (trait classes):

```{r permanova_tables_traits_plot, echo = FALSE}
options(knitr.kable.NA = '')  # ensure NAs printed as blanks

# ~ trait CWMs:
perm_li_traits_plot %>%
  mutate_at(3:5, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(6, 8), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Lichen functional trait composition vs. site')

perm_li_traits_plot_ph %>%
  mutate_at(2, ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(c(3, 5), ~ formatC(., digits = 3, format = "f")) %>%
  kable(caption = 'Post-hoc pairwise tests')
```

![Non-metric multidimensional scaling (MDS) plot of lichen community structure as (a) taxonomic and (b) functional groups, and (c) functional trait composition, using 'plot-level' data summarised for each plot. Based on Bray-Curtis dissimilarities (zero-adjusted for taxonomic and functional group abundances) calculated from $\log_{10}(x+1)$-transformed data. Ellipses represent the standard deviation of points within each group.](./figs/mds_plot.pdf)

Of the lichen taxonomic/functional groups making the greatest contributions to differences in community structure, [_R-Es_, _Gr-th_, _ArH_, _Gr-li_, _R-Py_, _PyP_, _CoL_, _CoC_ and _PiM_] and [_sqhya_, _crthe_, _crbya_, _crbysp_, _crlir_, _crpeb_, _sqhyp_, _focyb_, _focyr_, _crmzp_ and _cata_] exhibited higher mean abundances in old growth sites compared to the logged site, whereas [_ArC_, _RoBi_ and _UiCr_] and [_crstlf_, _crsco_ and _crapo_] showed the opposite trend.

SIMPER table (taxonomic groups):

```{r simper_table_taxa_plot, echo = FALSE}
# insert SIMPER tables:
# ~ taxonomic groups:
simp_li_taxa_plot[[1]] %>%
  rownames_to_column("Tax. group") %>%
  mutate_at(c(2:4), ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(5, ~ formatC(., digits = 1, format = "f")) %>%
  kable(
    caption = 'Lichen taxonomic group contributions to differences in community structure. (D,M), S: overall mean abundance in old growth and logged sites, respectively; Contrib./SD: average taxon contribution to group dissimilarity divided by standard deviation of contributions; Contrib.%: percent contribution of taxon to overall between-group dissimilarity. Only the most important taxa (Contrib. $\\geq$ 3%) are shown.'
  )
```

SIMPER table (functional groups):

```{r simper_table_func_plot, echo = FALSE}
# ~ functional groups:
simp_li_func_plot[[1]] %>%
  rownames_to_column("Func. group") %>%
  mutate_at(c(2:4), ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(5, ~ formatC(., digits = 1, format = "f")) %>%
  kable(
    caption = 'Lichen functional group contributions to differences in community structure'
  )
```

The following lichen functional traits exhibited greater CWM values in old growth sites compared to logged site: vegetative dispersal structures (i.e. soredia/isidia/pseudisidia), squamulose growth type, absence of fimbriate prothallus, non-byssoid surface structure, unknown Chlorophyta photobiont, apothecia with margins, and thelotremoid apothecia. Conversely, the following traits exhibited greater CWM values in the logged site: fimbriate prothallus, ?absence of sexual dispersal structures, byssoid surface structure, ?absence of vegetative dispersal structures, non-carbonised perithecia, crustose growth type, and Trentepohliaceae photobiont.

SIMPER table (trait classes):

```{r simper_table_traits_plot, echo = FALSE}
# ~ trait classes:
simp_li_traits_plot[[1]] %>%
  rownames_to_column("Trait class") %>%
  mutate_at(c(2:4), ~ formatC(., digits = 2, format = "f")) %>%
  mutate_at(5, ~ formatC(., digits = 1, format = "f")) %>%
  kable(
    caption = 'Lichen trait class contributions to differences in community structure. (D,M), S: overall mean CWM (community-level weighted mean value) in old growth and logged sites, respectively; Contrib./SD: average trait class contribution to group dissimilarity divided by standard deviation of contributions; Contrib.%: percent contribution of trait class to overall between-group dissimilarity. Only the most important trait classes (Contrib. $\\geq$ 3%) are shown.'
  )
```

Indicator Value analyses revealed a number of lichen taxonomic/functional groups and functional traits that were significantly associated with the combination of Danau and Maliau sites (i.e. old growth sites), rather than with each site separately; namely taxonomic groups [_Gr-th_, _Gr-li_, _PyP_ and _CoC_], functional groups [_crthe_, _crbya_, _crlir_, _crpeb_ and _focyr_] and traits [thelotremoid apothecia, lirellae, carbonised perithecia and Rhizonema photobiont]. Additionally, taxon [_ArC_] and functional group [_crstlf_] were significantly associated with the logged site.

Indicator Value analysis output (taxonomic groups):

```{r indval_summary_taxa_plot, echo = FALSE}
# insert Indicator Value summary output:
# ~ taxonomic groups:
cat(indval_taxa_plot_summ, sep = "\n")
```

Indicator Value analysis output (functional groups):

```{r indval_summary_func_plot, echo = FALSE}
# ~ functional groups:
cat(indval_func_plot_summ, sep = "\n")
```

Indicator Value analysis output (trait classes):

```{r indval_summary_traits_plot, echo = FALSE}
# ~ trait classes:
cat(indval_traits_plot_summ, sep = "\n")
```

The diversity of bark types and proportion of dipterocarps within plots were identified as key tree functional traits influencing lichen community structure, in terms of both taxonomic and functional groups (BIOENV; $\rho$ = `r rep_2dp(bioenv_taxa_plot_rho)` and $\rho$ = `r rep_2dp(bioenv_func_plot_rho)`, respectively) and lichen trait classes ($\rho$ = `r rep_2dp(bioenv_traits_plot_rho)`).

In all three cases, CAP (dbRDA) axis 1 correlated negatively with both plot-level tree traits, while axis 2 appeared to be related positively to bark diversity and negatively to dipterocarp proportion; the overall analysis was significant, as determined by permutation tests (taxonomic groups: P `r rep_p(cap_taxa_plot_p)`; functional groups: P `r rep_p(cap_func_plot_p)`; trait classes: P `r rep_p(cap_traits_plot_p)`).

however, while neither 'trait' was significant as a constraining variable for either lichen taxonomic groups or trait classes, only the proportion of dipterocarps was significantly related to lichen community structure in terms of functional groups (pseudo-_F_~1,19~ = `r rep_2dp(cap_func_plot_margin_f["dipterocarp_prop"])`, P `r rep_p(cap_func_plot_margin_p["dipterocarp_prop"])`).

Taxonomic groups [_ArA_, _ArT_, _CnS_, _CoC_, _Gr-th_, _PyP_, _R-Py_, _ThM_, _TryP_ and _VeFl_] and functional groups [_crbya_, _crbys_, _crlir_, _crmzs_, _crpeb_, _crspe_, _crthe_, _focyr_ and _sqhya_] appeared to be associated strongly with plot-level bark diversity. Taxa [_Gr-Li_, and _R-Es_] were also associated with bark diversity, but showed a stronger association with the proportion of dipterocarps. Conversely, groups [_ArC_ and _Uicr_] and [_crstlf_] were negatively associated with both bark diversity and proportion of dipterocarps.

In terms of lichen traits, carbonised perithecia, apothecia without margins and thelotremoid apothecia were associated with bark diversity and, to a lesser degree, the proportion of dipterocarps; unknown Chlorophyta photobiont, presence of hypothallus, squamulose growth type and vegetative dispersal structures were associated more strongly with proportion of dipterocarps. Crustose growth type, Trentepohliaceae photobiont, ?absence of vegetative dispersal structures and absence of hypothallus exhibited a strongly negative association with the proportion of dipterocarps.

![(a) CAP ordination plot of lichen community structure (taxonomic groups) according to continuous environmental (tree trait) variables, based on 'plot-level' data summarised for each plot. (b) Spearman rank correlation coefficients between CAP axes and individual taxa for which $|\rho| \geq$ `r cor_spp_x`.](./figs/cap_taxa_plot.pdf)

![(a) CAP ordination plot of lichen community structure (functional groups) according to continuous environmental (tree trait) variables, based on 'plot-level' data summarised for each plot. (b) Spearman rank correlation coefficients between CAP axes and individual functional groups for which $|\rho| \geq$ `r cor_spp_x`.](./figs/cap_func_plot.pdf)

![(a) CAP ordination plot of lichen functional trait composition according to continuous environmental (tree trait) variables, based on 'plot-level' data summarised for each plot. (b) Spearman rank correlation coefficients between CAP axes and individual functional groups for which $|\rho| \geq$ `r cor_spp_x`.](./figs/cap_traits_plot.pdf)

The overall success with which plots could be classified by site based on lichen community structure and trait composition (CAP) was `r rep_2dp(cap_taxa_plot_site$percent)`% (`r round(cap_taxa_plot_site$percent*22/100)`/22 plots) for taxonomic groups, `r rep_2dp(cap_func_plot_site$percent)`% (`r round(cap_func_plot_site$percent*22/100)`/22 plots) for functional groups and `r rep_2dp(cap_traits_plot_site$percent)`% (`r round(cap_traits_plot_site$percent*22/100)`/22 plots) for trait classes.

```{r cap_table_plot, echo = FALSE}
# insert CAP classification success summary table:
rbind(
  'Taxonomic groups' = cap_taxa_plot_site$percent.level,
  'Functional groups' = cap_func_plot_site$percent.level,
  'Trait classes' = cap_traits_plot_site$percent.level
) %>%
  as.data.frame %>%
  rownames_to_column('data') %>%
  mutate_at(c(2:4), ~ formatC(., digits = 1, format = "f")) %>%
  kable(caption = 'Percent classification success of plots to sites based on lichen taxonomic and functional group abundances and trait CWMs.')
```




## Queries

1. In both lichen abundance datasets, there are two _D810_ columns -- perhaps one of these (the first one) should be labelled _D710_, as this tree is not present in either dataset? This is what I've done. (Otherwise, should these be combined into a single column _D810_ in each case?)

1. I'm assuming that 'tree no.' is coded as follows: site-plot-tree (i.e. _D111_ = tree 11 in plot 1 in site Danum). Is this correct?
