---
title: 'How much taxonomic expertise is necessary for prioritising areas of lowland rainforest sites as habitats for lichens in northeast-Borneo?'
author: 'Rob Mrowicki'
date: '`r format(Sys.time(), "%d %B %Y")`'
output:
  pdf_document:
    fig_caption: yes
    fig_crop: yes
---

```{r setup, include = FALSE}
source('./scripts/00_setup.R')  # run setup script
```

```{r source_scripts, include = FALSE}
# run data scripts:
source('./scripts/01_import_tidy_and_manipulate_data.R')
source('./scripts/02_summarise_data.R')
source('./scripts/03_analyse_data.R')
```

## Field sites

Sampling was conducted at three sites within different areas of lowland forest in northeast Borneo, including two old growth reserve sites, (1) Danum Valley Conservation Area ('D') and (2) Maliau Conservation Area ('M'), and one selectively logged site, (3) the Stability of Altered Forest Ecosystems (SAFE) Project area ('S').

At each site, ...

The following functional traits were recorded for each tree: (1) girth, measured as ...; (2) bark type (S = smooth, C = scaly, R = ridged/rough, DR = dry ancient); and (3) the presence of buttresses or flutes.

## Data analysis

Taxonomic/functional group richness (_S_) and Shannon-Wiener diversity (H') were calculated for each tree.

Tree bark type was coded numerically in relation to increasing roughness, i.e. smooth = 0, scaly = 1, ridged/rough = 2, ancient dry = 3.

Community dissimilarities were calculated using the zero-adjusted Bray-Curtis coefficient (Clarke _et al._, 2006, _J. Exp. Mar. Biol. Ecol._ 330:55--80).

## Queries

1. In both lichen abundance datasets, there are two _D810_ columns -- is this a mistake (should these be combined into a single column in each case)? OR should one of these be labelled _D710_, as this tree is not present in either dataset?

1. The _crsco_ taxfree code is duplicated in the lichen functional group abundance dataset -- should the abundance variables be summed for these two rows?

1. There are no data for tree _SA4_ in the lichen functional group abundance dataset (in addition to tree _D710_ -- see above).

1. Can 'bark type' be coded numerically (i.e. an ordinal categorical variable), e.g. smooth = 1, scaly = 2, ridged/rough = 3, dry ancient = 4? In this way, it could be used as an environmental variable in the CAP analysis.
